Building Trust in Science Through a Bayesian Perspective on Probability and Uncertainty

### 2021-05-01 manuscript.tex
final version to submit

%%
%% This is file `./samples/longsample.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% apa7.dtx  (with options: `longsample')
%% ----------------------------------------------------------------------
%% 
%% apa7 - A LaTeX class for formatting documents in compliance with the
%% American Psychological Association's Publication Manual, 7th edition
%% 
%% Copyright (C) 2019 by Daniel A. Weiss <daniel.weiss.led at gmail.com>
%% 
%% This work may be distributed and/or modified under the
%% conditions of the LaTeX Project Public License (LPPL), either
%% version 1.3c of this license or (at your option) any later
%% version.  The latest version of this license is in the file:
%% 
%% http://www.latex-project.org/lppl.txt
%% 
%% Users may freely modify these files without permission, as long as the
%% copyright line and this statement are maintained intact.
%% 
%% This work is not endorsed by, affiliated with, or probably even known
%% by, the American Psychological Association.
%% 
%% ----------------------------------------------------------------------
%% 
\documentclass[man]{apa7}

@@

\abstract{
Uncertainty is ubiquitous in science and this uncertainty takes many forms, but the rhetoric around scientific knowledge and negotiating standards of evidence is often portrayed as immutable. Adding to this problem around the limited recognition of the tentative nature of science, when scientists try to communicate their uncertainty, research has shown us that many—including both learners and experts—struggle to interpret the statistics that are often used to make inferences from scientific data. We argue that a Bayesian approach can offer a different foundation and language around a core but sometimes unsettling tenet and persistent feature, uncertainty. Central to this conceptualization is viewing uncertainty probabilistically, as probabilities can provide a language for expressing the degree of uncertainty in our knowledge in a wide range of domains. In this essay, we (a) briefly introduce Bayes' theorem and wider ideas around Bayesian epistemology, (b) elaborate on the importance of probability and uncertainty in science and science education and review how people learn about these concepts and where they face difficulties, and (c) provide conceptual and practical examples within the JASP statistical software of how Bayes could help to address these issues. We conclude with implications and directions for future research focused around bolstering trust in science outside of classroom settings (for which Bayesian approaches could serve as a crosscutting part of science literacy) and in pre-collegiate science classrooms, where Bayes can provide a coherent approach to engaging in numerous science practices such as arguing from evidence and analyzing and interpreting data in a principled and flexible way.
}

\keywords{APA style, demonstration}

@@

Uncertainty is ubiquitous in science and this uncertainty takes many forms. For example, measurement uncertainty is present when there are differences (small or large) in the repeated measures of the same object or phenomenon \parencite{pls03} or when the object that is measured, like a plant, is inherently variable in its characteristics \parencite{pls03}. Another form of uncertainty is its role within explanatory theories, such as those associated with evolution and quantum physics, for which central models take the form of probablistic statements, i.e., uncertainty is a core constituent of these models \parencite{g00}.

Negotiating these many forms of uncertainty in argumentation—and building toward consensus at the same time that uncertainty is present—is a key part of the scientific process \parencite{t00}. In consequence, some philosophers of science have suggested that science is a process that builds increasingly better models which allows us to make increasingly more accurate predictions about the world  \parencite{g10, g06, n02, r77}.

However, while scientists and philosophers of science may be aware of the uncertainty (and limitations) surrounding scientific knowledge and negotiating standards of evidence interpretation, scientific knowledge is still often portrayed as immutable in educational contexts \parencite{d90}. In consequence, trust in science can erode when scientists individually or collectively change their views, such as during the first phase of the COVID-19 crisis in Germany, when scientists faced serious criticism from politicians and the media for changing their opinion based on new information. A similar tension was present in the United States and around the world regarding potential COVID-19 medication, when initial scientific data and changing interpretations of its utility collided with citizens’ skepticism (and, partisan political views).

Adding to this problem around the limited recognition of the tentative nature of science, when scientists try to communicate their uncertainty, research has shown us that many—including learners and experts alike—struggle to interpret statistics such as confidence intervals or error bars \parencite{gkv04, s07, tk74}. Similarly, difficulties in understanding the language of probability can contribute to a struggle on the part of students when learning about concepts such as evolution of nuclear decay that are modelled in terms of probabilistic statements \parencite{fth17}.

Based on accumulating evidence from cognitive psychology, neuroscience, and education, we build on and extend beyond the science and engineering practice of argumentation the foundational work of science educational \parencite{n11, s07} and statistics education scholars \parencite{a02, b02, h_a20, h_j20,me14, s07} to argue for a \EJ{I'd say ``underused'', because Bayes itself is pretty old and now pretty hip} approach to uncertainty and probability that is currently underrepresented in science education (research). Specifically, we argue that it is helpful to consider uncertainty in terms of a Bayesian epistemological framework to help people to understand uncertainty and probability better and to engage with them in a way that leverages their existing ways of thinking. Central to this conceptualization is viewing uncertainty probabilistically, as probabilities can provide a language for expressing the degree of uncertainty in our knowledge in a wide range of domains \parencite{g12}. In this way, we make this conjecture not only on the merits of Bayes' Theorem as a mathematical or statistical tool, but based on a body of psychological and educational research that shows that such an approach is an intuitive way for people to think about data and evidence in the world \parencite{g12} and in classroom contexts \parencite{ll20, n11, s07, so12, w17}.

The remainder of this paper is structured as follows. First, we will introduce Bayes theorem and wider ideas around Bayesian epistemology. Next, we will elaborate on the importance of probability and uncertainty in science (education) and review how people learn about these concepts and where they face difficulties. Then, we will provide examples of how Bayes could help to address these issues before concluding with implications and directions for future research.

\section{Primer on Bayes Theorem and Bayesian Views of Epistemology}

In essence, Bayes' theorem provides a framework for updating one's beliefs about the world from observations in a way that takes prior knowledge into account in a principled manner. Bayes' theorem (Equation 1) states that at any point in time, our knowledge about the world after making a new observation (i.e., our posterior knowledge) can be obtained by multiplying our prior knowledge with a predictive updating factor (e.g., \cite{RouderMorey2019,WagenmakersEtAl2016CD} ):
\begin{equation*}
        \text{Posterior uncertainty} = \text{Prior uncertainty} \times \text{Predictive updating factor}.
\end{equation*}
Mathematically, this can be expressed as  
\begin{equation}
\label{eq:BayesRule}
        p(\theta \mid \text{data}) = p(\theta) \times \frac{p(\text{data} \mid \theta)}{p(\text{data})},
\end{equation}
where $\theta$ is a hypothesis, proposition, claim, parameter value, or in general any particular account of the world about which we are uncertain. The predictive updating factor consists of two components. First, $p(\text{data} \given \theta)$ is the likelihood, that is, the extent to which the observed data were expected under $\theta$. Second, $p(\text{data})$ is the marginal likelihood, that is, the extent to which the observed data were expected without conditioning on $\theta$. If the observed data are more expected (i.e., better predicted) when conditioning on $\theta$, the predictive updating factor is higher than 1, and $\theta$ becomes more credible.

To make this more tangible, consider the following example. Suppose an educator is interested in the proportion of students that use Windows or MacOS as their operating system. The proportion of Windows users $\theta$ ranges between 0 and 1. Before you start our Bayesian analysis, you need to quantify your background knowledge -- the relative plausibility of the different values for $\theta$. Given their previous experiences as a teacher, their best guess is that about half of the students use MacOS. They would not be surprised if the proportion of Windows users was between 30\% and 70\% but you would be rather surprised to find either no Windows or exclusive Windows usage.

This background knowledge motivates the use of the dome-shaped prior distribution shown as the dashed line in Figure~\ref{fig:WindowsMacPriorPosterior}.\footnote{This is the beta$(2,2)$ distribution.} Now suppose you start to collect data by asking students about their operating systems. The first ten observations of Windows (W) and MacOS (M) usage that you make are $\{W, W, M, W, M, W, W, M, W, M\}$: six students use Windows, four use MacOS.

@@

Although we have demonstrated this updating process in a single step, it could also be executed sequentially, one student at a time. For instance, the first observation you make is a $W$ and in light of your prior this makes Windows usage a little more likely than MacOS usage. This slight change in knowledge is reflected in the difference between the distributions on the top two lines in Figure~\ref{fig:WindowsMacSequential}; line ``0'' represents the dome-shaped prior distribution and line ``1'' represents the posterior distribution after the first observations. Note that observing a $W$ has nudged the distribution a little to the right. Crucially, the `nudged' posterior distribution after the first observation now takes the role of the prior distribution ready to be updated by our second observation. The second observation is again a $W$, and line ``2'' shows that the resulting posterior distribution is nudged to the right once more. At every stage in the sequential updating process, the posterior distribution based on the observations seen so far becomes the prior distribution for incorporating the information from the next observation. Specifically, the changing distributions in Figure~\ref{fig:WindowsMacSequential} show our knowledge about the world is subject to constant change, by making observations and integrating incoming information with current knowledge using Bayes' theorem. It should be emphasized that although the updating process is conceptually different, the final outcome is identical: the shape of the posterior distribution is unaffected by whether data arrive simultaneously or sequentially.

@@

How do the principles from this example translate to students' work with data and evidence -- not only educators? A recent case from physics can serve to illustrate this approach in the context of students' activity in science classrooms. In 2011, physicists reported that they had observed elementary particles moving faster than the speed of light \parencite{b11} - and thus defying Einstein’s theory of relativity. However, instead of celebrating a historic discovery, the scientists started to look for what had gone wrong in their observations.

How can Bayes’ theorem help to make sense of this? With nearly a century of accumulated evidence that the theory of relativity holds, scientists had a very strong prior belief about the speed of light. To change this belief, extraordinary evidence would have been required. Thus, a single instance of apparently faster-than light elementary particles does not give a likelihood strong enough to change the prior in a meaningful way and in consequence the posterior, the knowledge about the world after making an observation, does not change in a meaningful way.

Thus, rather than changing their belief about the speed of light, scientists looked for things that could have gone wrong with the measurement. This example shows how a qualitative take on Bayes’ theorem can explain and predict many aspects of how animals \parencite{o13}, people \parencite{tgk06}, and organisations \parencite{kaj12} learn from observations. Recently, advances in computational power and techniques have started the more and more widespread application of this principle in a quantitative way that allows researchers to precisely express their knowledge about the world and then learn from observations.

Expressed in the Bayesian learning cycle (Figure \ref{fig:The Bayesian Learning Cycle}), a Bayesian reasoning framework allows scientists to integrate the strengths of both, inductive and deductive research approaches, in a formalized way. Further, when specifying the prior, scientists openly and explicitly need to describe the (un)certainty in the current state of knowledge and the posterior captures how the data has changed the knowledge and the (un)certainty in that knowledge. Thus, Bayesian reasoning is, at least in principle, open and transparent by design.

\begin{figure*}[h]
\begin{center}
\includegraphics[width = .65\paperwidth]{learning.png}
\caption{The Bayesian Learning cycle}
\label{fig:The Bayesian Learning Cycle}
\end{center}
\end{figure*}

Having introduced Bayes theorem and an application of Bayes theorem as a cycle through which individuals can learn, we describe the role of uncertainty in science, with a focus on the role of probability.

\section{Uncertainty in Science and the Role of Probability}

Probability and uncertainty are ubiquitous in science—in scientific methodologies, the scientific concepts, or in how science is communicated \textcite{gsoobmc17}. The practice of science often begins with measurement, and with measurement comes measurement error, which has a random component. This random component arises from random variations in the measurement process and limits the certainty that one can have in the measurements. Reducing that measurement error has been critical for many discoveries in science, such as in sorting out the periodic table as we know it today \textcite{fco15}. However, deciding on adequate statistical procedures to determine measurement uncertainty regularly sparks debate, e.g., the recent study on the relationship between SARS-CoV-2 viral load and patient age by \textcite{jmvbzhd20} was heavily debated in the scientific community \parencite{frick_peer-review_2020}.

Probability and uncertainty are not only key concepts in science practices such as analyzing and interpreting data, but also part of the scientific concepts themselves. One example is Heisenberg’s uncertainty principle or quantum mechanics more generally \textcite{feynman1951operator}. In quantum mechanics, confidence in measurement is limited not only by measurement error but in principle—even in the perfect experimental setup there is no deterministic outcome. Instead, one has to calculate the probability of each possible outcome using the Born rule. To this day, there is an ongoing argument among physicists and philosophers about how to interpret probability and the puzzling quantum mechanical phenomena such as quantum tunnelling  \textcite{c19}. Another example of how probability and uncertainty are central to learning scientific concepts is  evolution  \textcite{fsnh19}. Evolution is modelled as a probabilistic process as the emergence of new variations (e.g., through mutations) can only be described probabilistically. Further, the continued survival of these variations is governed by phenomena such as natural selection or random drift which also escape a deterministic description and are thus modelled probabilistically.\EJ{I agree that we can only describe the process probabilistically. But that does not mean it \emph{is} probabilistic.} \MK{tried to address this comment as well.} Similar to within quantum physics, how to interpret this probability has also sparked debate among researchers  \textcite{m16}.
Given that probability and uncertainty spark debate among scientists, it is perhaps unsurprising that misinterpretations happen when scientists in the natural and engineering sciences communicate about their findings with their peers or the public \parencite[]{gkv04, c14, mg17}. For example, error bars in the form of confidence intervals are routinely interpreted as distributions that assign higher probability to the center of the interval \parencite{kl18} or \emph{p}-values are interpreted as effect sizes \parencite{gc17, n14} (although p-values describe how (in)compatible a set of data is with a set of assumptions. In sum, misinterpretations of statements about probability and uncertainty are common across the range of modes in which scientists communicate about their work.

>>>>

%%
%% This is file `./samples/longsample.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% apa7.dtx  (with options: `longsample')
%% ----------------------------------------------------------------------
%% 
%% apa7 - A LaTeX class for formatting documents in compliance with the
%% American Psychological Association's Publication Manual, 7th edition
%% 
%% Copyright (C) 2019 by Daniel A. Weiss <daniel.weiss.led at gmail.com>
%% 
%% This work may be distributed and/or modified under the
%% conditions of the LaTeX Project Public License (LPPL), either
%% version 1.3c of this license or (at your option) any later
%% version.  The latest version of this license is in the file:
%% 
%% http://www.latex-project.org/lppl.txt
%% 
%% Users may freely modify these files without permission, as long as the
%% copyright line and this statement are maintained intact.
%% 
%% This work is not endorsed by, affiliated with, or probably even known
%% by, the American Psychological Association.
%% 
%% ----------------------------------------------------------------------
%% 
\documentclass[man]{apa7}
% \documentclass[man, mask, floatsintext]{apa7}

@@

\abstract{
Uncertainty is ubiquitous in science, but scientific understanding is often represented to the public and in educational contexts as certain and immutable. Adding to this problem is that when scientists try to communicate their uncertainty, research has demonstrated that many people--learners and experts alike--struggle to interpret the statistics that are often used to make inferences about evidence from scientific data. We conjecture that a Bayesian perspective can offer a foundation for teaching and learning about a persistent but sometimes unsettling tension for science, making sense of and communicating about uncertainty. Central to this approach is viewing uncertainty \ital{probabilistically}, as doing so provides a language for expressing the degree of uncertainty in knowledge across a range of scientific domains and socio-scientific issues. In this paper, we elaborate on the importance of probability and uncertainty in science (and science education) and review how people learn about these concepts---and when and why they face difficulties when doing so. Then, we provide an accessible primer on Bayes Theorem and more general principles following from the application of Bayes Theorem. Next, we provide examples of how a Bayesian perspective could inform efforts around both science literacy and science education with an emphasis upon building trust in science. We conclude with directions for future research focused on bolstering trust in science outside of and within pre-collegiate science classrooms, where a Bayesian perspective can provide a coherent approach to engaging in science to make sense of scientific ideas in a principled yet flexible manner.
}

\keywords{Bayesian, uncertainty, probability, trustworthiness, statistics}

@@

Uncertainty is ubiquitous in science and creating ways to account for this uncertainty has helped science to progress. Consider, for instance, uncertainty due to the measurements or instruments used to study phenomena: In such cases, error due to measurement can be incorporated into models through the repeated measures of the same object or phenomenon \parencite{pls03}. Measurement is also present when the object that is measured--like a plant--is variable in its characteristics because of environmental and genetic variation (and interactions among these sources of variation) between the individual plants \parencite{pls03}. Another form of uncertainty is accounted for in terms of its central role within explanatory theories--such as those associated with evolution and quantum physics--for which models take the form of probabilistic statements about the world \parencite{g00}.

Negotiating the many forms of uncertainty through (constructively) arguing—-and building toward consensus at the same time that uncertainty is present—-is a key part of the scientific process \parencite{duschl2008science, t00, nrc12}. Concomitantly, some philosophers of science have suggested that science is a process that builds increasingly better models which allows us to make increasingly more accurate predictions \parencite{g10, g06, n02, r77}. However, while scientists and historians and philosophers of science are well aware of the uncertainty (and limitations) around scientific knowledge and the practice of negotiating and arguing from evidence, scientific knowledge is often represented to the public and in educational contexts as certain and immutable \parencite{d90}. Trust in science can erode and has eroded when scientists individually or collectively change their views, such as during the first phase of the COVID-19 crisis, when scientists faced serious criticism from the public, the media, and elected representatives for changing their recommendations based upon new scientific data and findings. A similar tension is now present concerning vaccines for COVID-19--a tension that involves reconciling robust but still initial clinical trial data and new data from population-level vaccination efforts and making policy-level recommendations (and individual decisions) amid this uncertainty.
Adding to this difficult state of affairs regarding the role of uncertainty in science is that when scientists try to communicate their uncertainty, prior research has demonstrated that many people–learners and experts alike–struggle to interpret the statistics and derivations of statistics (such as standard errors/margins of error and confidence intervals) and that are often used to make inferences from scientific data \parencite{gkv04, s07, tk74}. Similarly, difficulties in understanding the technical language of probability can contribute to a struggle on the part of students when learning about concepts such as evolution and nuclear decay that are modeled in terms of probabilistic statements \parencite{fth17}.

In this paper, we argue that these challenges concerning trust in scientific knowledge that not only can but must change can be addressed in a principled, flexible way, but that doing so requires making use of a Bayesian perspective that heretofore has been under-used outside of particular scientific and engineering domains---domains that often make little contact with science educators and researchers with an interest in building trust in science. Yet, the opportunity to do so now exists for technical and social reasons. Based on accumulating evidence from psychologists using Bayesian models of cognition \parencite{g12, tgk06}, statisticians and statistics educators advancing tools and practices for Bayesian methods \parencite{a02, b02, h_a20, h_j20,me14, s07} and science education scholars who have advanced a Bayesian perspective on argumentation \parencite{n11, s07}, we conjecture that a Bayesian perspective can help people--and science learners--to understand and reason about uncertainty in a way that leverages their existing ways of thinking. Central to this conceptualization is viewing uncertainty \ital{probabilistically}, as probabilities can provide a language for expressing the degree of uncertainty in our knowledge in a wide range of domains \parencite{g12}. In drawing on research from psychologists, learning scientists, and educational researchers, we make this conjecture not only on the merits of Bayes' Theorem as a mathematical or statistical tool, but primarily on evidence that the perspective we are advocating can provide an intuitive way for people to think about data and evidence in the world \parencite{g12} and in classroom contexts \parencite{ll20, n11, s07, so12, w17}.

\section{Uncertainty in Science and the Central Role of Probability}

Probability and uncertainty are ubiquitous in scientific methodologies, scientific concepts, and in how science is communicated \textcite{gsoobmc17}. The practice of science often begins with measurement for which measurement error--which has a random component--is endemic. This random component arises from random variations in the measurement process and limits the certainty that one can have in the measurements. Reducing that measurement error has been critical for many discoveries in science, such as in sorting out the periodic table as we know it today \parencite{fco15}. However, deciding on adequate statistical procedures to determine measurement uncertainty regularly sparks debate, e.g., the recent study on the relationship between SARS-CoV-2 viral load and patient age by \textcite{jmvbzhd20} was heavily debated in the scientific community \parencite{frick_peer-review_2020}.

Probability and uncertainty are not only key concepts in science practices (such as analyzing and interpreting data) but are also an integral part of explanatory models of and theories for scientific concepts. For example, in the context of Heisenberg’s uncertainty principle and--more generally--quantum mechanics \parencite{feynman1951operator}, confidence is limited not only by measurement error but in principle---even in the perfect experimental setup there is no deterministic outcome. Instead, one has to calculate the probability of each possible outcome using the Born rule. 
To this day, there is an ongoing argument among physicists and philosophers about how to interpret probability and the puzzling quantum mechanical phenomena such as quantum tunneling \parencite{c19}. Another example of how probability and uncertainty are central to learning scientific concepts is evolution \parencite{fsnh19}. Evolution is modeled as a probabilistic process as the emergence of new variations (e.g., through mutations) can only be described probabilistically. Further, the continued survival of these variations is governed by phenomena such as natural selection or random drift which also escape a deterministic description and are thus modeled probabilistically. Similar to within quantum physics, how to interpret this probability has also sparked debate among researchers \parencite{m16}. 
Given that probability and uncertainty spark debate among scientists, it is perhaps unsurprising that misinterpretations happen when scientists in the natural and engineering sciences communicate about their findings with their peers or the public \parencite[]{gkv04, c14, mg17}. For example, error bars in the form of confidence intervals are routinely interpreted as distributions that assign a higher probability to the center of the interval \parencite{kl18} or \emph{p}-values are interpreted as effect sizes \parencite{gc17, n14} (although p-values describe how (in)compatible a set of data is with a set of assumptions. In sum, misinterpretations of statements about probability and uncertainty are common across the range of modes in which scientists communicate about their work.

\subsection{The Roles of Probability Across Scientific Disciplines}

Given the above-noted roles of probability in scientific concepts and theories, it is unsurprising that educational researchers have documented how students learn about probability. As one example, probability and chance events play an important role in the life sciences \parencite{g03, gk08}, particularly in evolutionary processes and learning about them \parencite{th17}. Recently, \textcite{fsnh19} investigated the relation between statistical reasoning and acceptance and knowledge about evolution in a large sample of nearly 500 university students in the US. They found that students’ statistical reasoning capabilities are a strong predictor of both acceptance of evolution and knowledge about evolution.

In the physical sciences, students' struggles with the probabilistic nature of quantum physics \parencite{br02} and nuclear decay \parencite{smmv17} have been established, but the underlying reasons and mechanisms remain less researched than in evolution. Research into students’ misconceptions inspired by conceptual change research has started to address this gap \parencite{ms15, alra14, st09}. Findings mirror the position of \textcite{bgv94}. Specifically, the concepts of probability are taught abstractly and in such a way that does not build upon the ideas that students already hold, the experiences that students have, and the language that students use. Thus, these studies affirm students' struggles with scientific concepts that are probabilistic but hint at an underlying issue of more foundational challenges with reasoning about probability beyond science education contexts.

Indeed, past research has shown that the language and representations used to communicate probability and uncertainty independent of specific scientific content domains or scientific practices are also known to present difficulties to learners. The work of \textcite{tk74}, especially, points to a whole range of biases that the language in which statements about probability and uncertainty are framed can elicit. A prominent example is the neglect of base rate information in the (timely) context of medical testing, that is, the probability of having a medical condition based on a positive test result is often hugely overestimated because information about the base rate of people affected from the condition is neglected \parencite{kahneman_thinking_2012}.

Responding to and building upon Kahneman and Tversky's work, Gigerenzer and colleagues provided important insights into the underlying mechanisms for the occurrence of these biases. Their research suggests that the biases are at least partly the result of information about probability and uncertainty being provided in a way that is incompatible with the heuristics that people develop from their everyday experiences \parencite{gh95, jkg18}. Based upon this premise, \textcite{jkg18} were able to demonstrate that students judge probability significantly better when language is adopted that better aligns with the heuristics that most learners develop from their everyday experiences. In sum, it seems feasible to change representations of probability and uncertainty so that they better align with ideas and heuristics that learners hold so that these support, rather than hinder, learning.

\subsection{Uncertainty in the Context of Analyzing and Interpreting Data}

Recent work--primarily by developmental psychologists and scientists--has applied Bayesian ideas to a new context: how children make sense of and world \parencite{g12, tgk06, tgk06}. This notion that children reason about the world in a way that is analogous to Bayes Theorem is in some ways surprising, but it also builds upon much earlier work that also documented the surprisingly complex ways that children could reason about the world---as well as the assumed limitations of children's reasoning, the work of \textcite{pi69}. Different from earlier work on the stages through which children develop, recent work using Bayesian models considers the differences in the reasoning ability of children and adults to be a matter of degree, rather than kind. In short, children hold initial ideas and understandings that they change (or update) in ways that are concordant with how Bayes Theorem represents the updating of an initial belief in light of data (and how the Bayesian Learning Cycle described earlier allows individuals to integrate inductive and deductive reasoning).

In addition to developmental research that uses Bayes Theorem as an analog for how children make sense of data and information in light of their initial ideas and beliefs, other work bears a striking semblance to developmental research that does. In particular, research on how learners’ prior understanding influence how they make sense of how likely different patterns in data are centers the same tension around interpreting data in light of initial beliefs that a Bayesian perspective highlights \parencite[e.g., ][]{kd88, mkm07, ssc07}. While these studies have focused on how students’ prior ideas and data interact, Bayesian frameworks were not used to formalize this process of integrating data.

A Bayesian view of how children learn from their experiences and update their beliefs about the world has become especially prominent over the past 10-15 years \parencite{gt07, gw12}. Because of this, one might assume that there has been a ready application of these ideas about how children think about the world in a Bayesian way to education and developmental scientists have pointed to this opportunity. For instance, \textcite{g12} wrote that the use of Bayesian methods to understand child development can serve as “a scientific foundation for a long tradition on ‘inquiry-based’ science education,” that “could lead us to much more specific and scientifically supported proposals for education” (p. 1627).

However, despite arguments such as Gopnik's (2012) for integrating developmental and educational research, it has largely not been the case that “science itself could help turn young children’s natural curiosity and brilliance into better science teaching and learning” \textcite[p. 1627]{g12}, as some science education scholars have bemoaned \parencite{ls15}). Indeed, scholars have argued that it is often obsolete applications of Piaget’s ideas \parencite[e.g., ][]{pi69} that represent the greatest contribution of developmental ideas to education \textcite{ls07}. We argue that this is still largely the case in the present, despite advances in both developmental research as well as statistical methodology related to Bayesian ideas—and that it is time to start exploring how bringing out the inherent Bayesian ways of thinking in education may provide support for learners’ efforts to make sense of the world.

At the same time that opportunities to connect developmental science with education (especially science education), science education research has emphasized how important students' prior knowledge and understandings are to their understanding of scientific concepts and processes \parencite[e.g., ][]{ssc07}. For example, \textcite{mkm07} summarized findings across several studies showing that when elementary grade students reasoned with data in a context about which they had strong and valid prior understandings that informed their interpretation of data (about force and motion in the context of ramps), and they were able to explain sources of error in the data \parencite{mk03}, but when the data was in a context they did not understand well, they struggled to distinguish between random variation due to sources of error and sources of variation due to something that was causing a difference in the data (generated by measuring the motion of pendulums). Later design- and laboratory-based research has shown that students can develop an understanding of scientific ideas through the analysis of data \parencite{ls04,mkk17}.

In summary, developmental and psychological research into Bayesian models of cognition supports and bolsters efforts in education--especially science education--to design instruction based around eliciting and understanding students’ ideas \parencite{gb16, hbavb20, wtbs12}. There is an opportunity to use Bayesian ideas as a framework for formalizing these ideas about how prior ideas affect students’ reasoning about data and how teachers can leverage students’ ideas to design instruction, but a Bayesian perspective has--with a few exceptions focusing on how Bayes Theorem can provide a model for the science practice of argumentation \parencite[e.g., ][]{n11, so12}--has been absent from research in science education.

After having reviewed how people learn about probability and uncertainty and where they face difficulties and research on uncertainty in the context of analyzing data, we next formally introduce Bayes Theorem and describe the broader process through which one's beliefs can be revised, or updated, in a way that leverages both inductive reasoning from data and deductive reasoning from beliefs or principles--what we refer to as the \ital{Bayesian Learning Cycle}. We also describe some of the ways of knowing, or epistemological considerations, that follow from Bayes Theorem and which comprise a core part of the Bayesian perspective that we argue has a role in building trust in science.

\section{A Primer on Bayes Theorem and a Bayesian Epistemology}

\subsection{An Accessible Introduction to Bayes Theorem}

At its core, Bayes' theorem provides a means to update one's beliefs about the world from observations in such a way that prior knowledge can be used in a principled manner. As represented in Equation 1, Bayes' Theorem states that at any point in time, our knowledge about the world after making a new observation (i.e., our posterior knowledge) can be obtained by multiplying our prior knowledge with a predictive updating factor (e.g., \cite{RouderMorey2019,WagenmakersEtAl2016CD} ):
\begin{equation*}
        \text{Posterior uncertainty} = \text{Prior uncertainty} \times \text{Predictive updating factor}.
\end{equation*}
Mathematically, this can be expressed as  
\begin{equation}
\label{eq:BayesRule}
        p(\theta \mid \text{data}) = p(\theta) \times \frac{p(\text{data} \mid \theta)}{p(\text{data})},
\end{equation}
where $\theta$ is a hypothesis, proposition, claim, parameter value, or in general any particular account of the world about which we are uncertain. The predictive updating factor consists of two components. First, $p(\text{data} \given \theta)$ is the likelihood, that is, the extent to which the observed data were expected under $\theta$. Second, $p(\text{data}$ is the marginal likelihood, that is, the extent to which the observed data were expected without conditioning on $\theta$. If the observed data are more expected (i.e., better predicted) when conditioning on $\theta$, the predictive updating factor is higher than 1, and $\theta$ becomes more credible.

To illustrate the above equation, consider the following: An educator is interested in the proportion of students that use Windows or MacOS as their operating system. The proportion of Windows users $\theta$ ranges between 0 and 1. Before you start our Bayesian analysis, you need to quantify your background knowledge -- the relative plausibility of the different values for $\theta$. Given their previous experiences as a teacher, their best guess is that about half of the students use Windows. They would not be surprised if the proportion of Windows users was between 30\% and 70\% but you would be rather surprised to find either no Windows or exclusive Windows usage.

This background knowledge motivates the use of the dome-shaped prior distribution shown as the dashed line in Figure~\ref{fig:WindowsMacPriorPosterior}.\footnote{This is the beta$(2,2)$ distribution.} Now suppose you start to collect data by asking students about the operating systems of their computers. The first ten observations of Windows (W) and MacOS (M) usage that you make are $\{W, W, M, W, M, W, W, M, W, M\}$: six students use Windows and four use MacOS.

@@

Although we demonstrated this updating process in a single step, it could also be executed sequentially, one student at a time. For instance, the first observation you make is a $W$ and in light of your prior this makes Windows usage a little more likely than MacOS usage. This slight change in knowledge is reflected in the difference between the distributions on the top two lines in Figure~\ref{fig:WindowsMacSequential}; line ``0'' represents the dome-shaped prior distribution and line ``1'' represents the posterior distribution after the first observations. Note that observing a $W$ has nudged the distribution a little to the right. Crucially, the `nudged' posterior distribution after the first observation now takes the role of the prior distribution ready to be updated by our second observation. The second observation is again a $W$, and line ``2'' shows that the resulting posterior distribution is nudged to the right once more. At every stage in the sequential updating process, the posterior distribution based on the observations seen so far becomes the prior distribution for incorporating the information from the next observation. Specifically, the changing distributions in Figure~\ref{fig:WindowsMacSequential} show our knowledge about the world is subject to constant change, by making observations and integrating incoming information with current knowledge using Bayes' theorem. It should be emphasized that although the updating process is conceptually different, the final outcome is identical: the shape of the posterior distribution is unaffected by whether data arrive simultaneously or sequentially.

@@

\subsection{Integrating Inductive and Deductive Reasoning With the Bayesian Learning Cycle}

How do the principles from the above example translate to scientific data? A recent case from the physical sciences can serve to illustrate this approach. In 2011, physicists reported that they had observed elementary particles moving faster than the speed of light \parencite{b11} - and thus defying Einstein’s theory of relativity. However, instead of celebrating a historic discovery, the scientists started to look for what had gone wrong in their observations. With nearly a century of accumulated evidence that the theory of relativity holds, from a Bayesian perspective, scientists had a very strong prior belief about the speed of light. To change this belief, extraordinary evidence would have been required. Thus, a single instance of apparently faster-than-light elementary particles does not give a likelihood strong enough to change the prior in a meaningful way, and in consequence, the posterior--the knowledge about the world after making an observation--does not change in a meaningful way.
Thus, rather than changing their belief about the speed of light, scientists looked for things that could have gone wrong with the measurement. This example shows how a qualitative take on Bayes’ theorem can explain and predict many aspects of how animals \parencite{o13}, people \parencite{tgk06}, and organizations \parencite{kaj12} learn from observations. Recently, advances in computational power and techniques have started the more and more widespread application of this principle in a quantitative way that allows researchers to precisely express their knowledge about the world and then learn from observations. 
Expressed in the Bayesian learning cycle (Figure \ref{fig:The Bayesian Learning Cycle}), a Bayesian reasoning framework allows scientists to integrate the strengths of both, inductive and deductive research approaches, in a formalized way. Further, when specifying the prior, scientists openly and explicitly need to describe the uncertainty in the current state of knowledge, and the posterior captures how the data has changed the knowledge and the uncertainty in that knowledge. Thus, Bayesian reasoning is, at least in principle, open and transparent by design.
\begin{figure*}[h]
\begin{center}
\includegraphics[width = .65\paperwidth]{learning.png}
\caption{The Bayesian learning cycle is a reasoning framework that allows scientists, citizens, and learners to integrate the strengths of inductive and deductive reasoning while making sense of data in light of prior beliefs.}
\label{fig:The Bayesian Learning Cycle}
\end{center}
\end{figure*}
Having introduced Bayes Theorem and an application of Bayes Theorem as a cycle through which individuals can learn, we show how Bayes Theorem can be used to demonstrate several principles that are relevant to building trust in science.

\section{The Epistemological Principles that Follow from Bayes Theorem}

In the previous section, we emphasized the idea that hypotheses, or ideas, gain credibility when they predict the data well, and lose credibility when they predict new data poorly \parencite{WagenmakersEtAl2016CD}. This idea generalizes to the ``Fundamental Inductive Pattern'':
\begin{quotation}
``This inductive pattern says nothing surprising. On the contrary, it expresses a belief which no reasonable person seems to doubt: \emph{The verification of a consequence renders a conjecture more credible}. With a little attention, we can observe countless reasonings in everyday life, in the law courts, in science, etc., which appear to confirm to our pattern.'' \parencite[pp. 4-5]{Polya1954Vol2}
\end{quotation}

The rule from Equation~\ref{eq:BayesRule} includes the constant term $p(\text{data})$, which does not involve $\theta$. As a result, Bayes' rule can be re-written as:
\begin{equation}
\label{eq:BayesRulePropto}
p(\theta \mid \text{data}) \propto p(\text{data} \mid \theta) \times p(\theta),
\end{equation}
where `$\propto$' stands for `is proportional to'. Thus, Bayes' rule states that our posterior knowledge $p(\theta \given \text{data})$ is proportional to the likelihood $p(\text{data} \given \theta)$ (or the extent to which the observed data are expected given $\theta$) multiplied by our prior knowledge $p(\theta)$.

Equation~\ref{eq:BayesRulePropto} emphasizes that the posterior uncertainty about $\theta$ is a \emph{compromise} between our prior uncertainty about $\theta$ and the predictive performance of $\theta$. But, the posterior uncertainty after having observed the first datum becomes the prior uncertainty for the next datum---and any other data (cf. Figure~\ref{fig:WindowsMacSequential}). Consequently, after having observed another datum, the posterior uncertainty represents a compromise of a compromise. As the data accumulate, the posterior uncertainty is more and more determined by predictive performance, and the impact of the initial uncertainty about $\theta$ is increasingly watered down: `the data overwhelm the prior' \parencite{WrinchJeffreys1919}.

This implies (but does not dictate, as we discuss next) that a constant stream of accumulating data ought to bring any two people into an arbitrarily close agreement, no matter how divergent their opinions may have been at the outset. There are two important caveats here.
